Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-10-10 paper/paper.tex
reduce to page limit

Due to space constraints, we can offer only a brief overview of multiparameter
persistent homology. For a primer on persistent homology,
see~\cite{ghrist_barcodes:_2008, carlsson_topology_2009}; for multiparameter
persistent homology, see~\cite{lesnick_interactive_2015}. An introduction to
lattices may be found in~\cite{davey2002introduction}.

@@

This representation theoretic fact gives a representation\footnote{Pun intended.} of
$\R$-indexed persistence modules via \introduce{barcodes} or
\introduce{persistence diagrams}. Each bar $I_{[a,b)}$ in the barcode represents
a homology class which is \introduce{born} at $a$ and \introduce{dies} at $b$.

@@

While there does not exist a discrete set of invariants for $M$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: R_+ \times R \to \Z_+;\quad \mathbf{x} \mapsto \dim(M_\mathbf{x}),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: L \to \Z_+$, for $j =
0,1,2$. For $M = PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about locations of births
and deaths of persistence classes.

@@

A \introduce{lattice} $L$ is a partially ordered set in which every pair of
elements $x,y$ has a greatest lower bound (the \introduce{meet} $x \wedge y$)
and a least upper bound (the \introduce{join} $x \vee y$). These operations
and their properties produce an algebraic characterization of lattices; the
order theoretic and algebraic properties of a lattice can be recovered from each
other. The key insight of~\cite{pueschel_discrete_2019} is that the meet and
join operations on a lattice define two ``shift operators'' that can be
exploited to define convolutional filters for signals on a lattice.

@@

A particularly nice class of examples of lattices are given by the sets $\R^n$
viewed as partially ordered sets, with the ordering $(x_1,\ldots,x_n) \leq
(y_1,\ldots,y_n)$ whenever $x_i \leq y_i$ for all $1 \leq i \leq n$. These are,
of course, the indexing sets for persistent homology, suggesting that lattice
convolutions over $\R^n$ or its finite sublattices may be useful in analyzing
data coming from multiparameter persistence computations.

@@

\section{Experiments}
\begin{figure}[b]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
    \includegraphics[width=0.49\textwidth]{lattice_accuracy.png}
    \vspace{0.2cm}
    \includegraphics[width=0.49\textwidth]{conv_accuracy.png}
    \end{center}

@@

As the filter function on these data sets, we use
\[\rho_{\text{codense}}(x;k) = \frac{1}{\frac{1}{k} \sum_{y \in N_k(x)}
    d(x,y)},\]
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$. This is the
\introduce{codensity} filtration, so named because the points in the densest
regions of $\mathcal M$ will appear first.

@@

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of noisy training as well as dramatically increased training time.  
%% more discussion of what the data actually looks like

\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.

We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?

>>>>

Space constraints require that this section be laconic. For a primer on
persistent homology, see~\cite{ghrist_barcodes:_2008, carlsson_topology_2009};
for multiparameter persistent homology, see~\cite{lesnick_interactive_2015}. An
introduction to lattices may be found in~\cite{davey2002introduction}.

@@

% This representation theoretic fact gives a representation\footnote{Pun intended.} of
% $\R$-indexed persistence modules via \introduce{barcodes} or
% \introduce{persistence diagrams}. Each bar $I_{[a,b)}$ in the barcode represents
% a homology class which is \introduce{born} at $a$ and \introduce{dies} at $b$.

@@

While there does not exist a discrete set of invariants for $M$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: R_+ \times R \to \Z_+;\quad \mathbf{x} \mapsto \dim(M_\mathbf{x}),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: L \to \Z_+$, for $j =
0,1,2$. For $M = PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about births
and deaths of homology classes.

@@

A \introduce{lattice} $L$ is a partially ordered set in which every pair of
elements $x,y$ has a greatest lower bound (the \introduce{meet} $x \wedge y$)
and a least upper bound (the \introduce{join} $x \vee y$). These operations
and their properties produce an algebraic characterization of lattices; the
ordering can be recovered from the algebra and vice versa. The key insight of~\cite{pueschel_discrete_2019} is that the meet and
join operations on a lattice define two ``shift operators'' that can be
exploited to define convolutional filters for signals on a lattice.

@@

A nice class of examples of lattices are given by the sets $\R^n$
viewed as partially ordered sets, with the ordering $(x_1,\ldots,x_n) \leq
(y_1,\ldots,y_n)$ whenever $x_i \leq y_i$ for all $1 \leq i \leq n$. These are,
of course, the indexing sets for persistent homology, suggesting that lattice
convolutions over $\R^n$ or its finite sublattices may be useful in processing
data coming from multiparameter persistence computations.

@@

\section{Experiments}
\begin{figure}[b]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network. Results are averaged over 10 runs.}\label{fig:comparison}
    \begin{center}
    \includegraphics[width=0.49\textwidth]{lattice_accuracy.png}
    \vspace{0.2cm}
    \includegraphics[width=0.49\textwidth]{conv_accuracy.png}
    \end{center}

@@

As the filter function on these data sets, we use the \introduce{codensity} function
\[\rho_{\text{codense}}(x;k) = \frac{1}{\frac{1}{k} \sum_{y \in N_k(x)}
    d(x,y)},\]
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$.

% This is the
% \introduce{codensity} filtration, so named because the points in the densest
% regions of $\mathcal M$ will appear first.

@@

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of more variability in performance during
training. %noisy training as well as dramatically increased training time.  
%% more discussion of what the data actually looks like

\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?
